---
title: "P8108_final_project_p3"
author: "Zihan Wu"
date: "2024-12-04"
output: html_document
---

```{r setup, include=FALSE, waring = F}
knitr::opts_chunk$set(echo = TRUE)
library(survival)
library(survminer)
library(tidyverse)
library(rms)
library(car)
library(MASS)
heart_failure_data <- read_csv("heart_failure_clinical_records_dataset.csv")
```
```{r}
# initial cox model
cox_full <- coxph(Surv(time, DEATH_EVENT) ~ age + anaemia + creatinine_phosphokinase + diabetes + ejection_fraction + high_blood_pressure + platelets + serum_creatinine + serum_sodium + sex + smoking, data = heart_failure_data)

# check non-linear relationships
par(mfrow=c(2,3))
continuous_vars <- c("age", "creatinine_phosphokinase", "ejection_fraction", 
                    "platelets", "serum_creatinine", "serum_sodium")

for(var in continuous_vars) {
    mart_res <- residuals(cox_full, type="martingale")
    plot(heart_failure_data[[var]], mart_res,
         xlab=var, ylab="Martingale Residuals",
         main=paste("Residuals vs", var))
    lines(lowess(heart_failure_data[[var]], mart_res), col="red")
}
```

Based on the Martingale residual plots analysis, we found non-linear relationships with the outcome in four continuous variables: creatinine_phosphokinase, and serum_creatinine, suggesting these variables need transformations for the Cox model.

### Cox

```{r}
# transform non-linear
heart_failure_data <- heart_failure_data %>%
  mutate(
    log_cp = log(creatinine_phosphokinase),
    log_serum_creatinine = log(serum_creatinine)
  )
# initial model with transformed and linear variables
cox_full <- coxph(
  Surv(time, DEATH_EVENT) ~ age + log_cp + ejection_fraction + 
    platelets + log_serum_creatinine + serum_sodium +
    anaemia + diabetes + high_blood_pressure + sex + smoking,
  data = heart_failure_data
)
#backward stepwise
cox_backward <- step(cox_full, direction = "backward")
#forward stepwise
cox_null <- coxph(Surv(time, DEATH_EVENT) ~ 1, data = heart_failure_data)
cox_forward <- step(cox_null, 
                   scope = ~ age + log_cp + ejection_fraction + 
                     platelets + log_serum_creatinine + serum_sodium +
                     anaemia + diabetes + high_blood_pressure + sex + smoking,
                   direction = "forward")

# Perform stepwise selection using stepAIC
cox_step  <- stepAIC(cox_full, 
                     direction = "both",    
                     trace = TRUE)          

# View final model
summary(cox_step )
# interactions
cox_interaction <- coxph(
  Surv(time, DEATH_EVENT) ~ age + log_cp + ejection_fraction + 
    log_serum_creatinine + serum_sodium +
    age:ejection_fraction + age:log_serum_creatinine,
  data = heart_failure_data
)
# compare
aic_comparison <- AIC(cox_full, cox_backward, cox_forward, cox_interaction)
print(aic_comparison)
# cox_forward & backward least AIC
# all age ejection_fraction log_serum_creatinine anaemia high_blood_pressure
# model's variable significance
final_model <- cox_forward
summary(final_model)

# VIF test for final selected variables
vif(lm(time ~ age + ejection_fraction + log_serum_creatinine + 
       anaemia + high_blood_pressure,
       data = heart_failure_data))
```
All VIF values are close to 1 (range: 1.01-1.08)
No multicollinearity issues
age ejection_fraction log_serum_creatinine anaemia high_blood_pressure

### AFT
```{r}
# Weibull AFT model
weib_full <- survreg(
  Surv(time, DEATH_EVENT) ~ age + log_cp + ejection_fraction + 
    platelets + log_serum_creatinine + serum_sodium +
    anaemia + diabetes + high_blood_pressure + sex + smoking,
  dist = "weibull",
  data = heart_failure_data
)
weib_step <- stepAIC(weib_full, direction = "both")

# Exponential AFT model
exp_full <- survreg(
  Surv(time, DEATH_EVENT) ~ age + log_cp + ejection_fraction + 
    platelets + log_serum_creatinine + serum_sodium +
    anaemia + diabetes + high_blood_pressure + sex + smoking,
  dist = "exponential",
  data = heart_failure_data
)
exp_step <- stepAIC(exp_full, direction = "both")
summary(weib_step)
summary(exp_step)
AIC(weib_step, exp_step, cox_step)
```
cox: age ejection_fraction log_serum_creatinine anaemia high_blood_pressure